---
title: "Assignment 04: Tidy Data and Joins"
author: "Sihaoyu Gao"
output:
  html_document:
    toc: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load, warning = FALSE, echo = FALSE, message=FALSE}
library(gapminder)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tidyverse)
```

## Exercise 1: Univariate Option 1

### 1. Wider Format

Make a tibble with one row per year, and columns for life expectancy for China and Canada.

```{r}
gapminder1 <- gapminder %>%
  filter(country %in% c("China","Canada")) %>% 
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = lifeExp,
              names_prefix = "Life Expectancy for ") 

DT::datatable(gapminder1)
```

### 2. Plot

The plot shows the scatterplot of life expectancy for China against Canada, with year beside each point.

```{r,fig.pos="center"}
gapminder1 %>%
  ggplot(aes(`Life Expectancy for China`, `Life Expectancy for Canada`)) +
  geom_point() +
  geom_text(aes(label = year), size = 3, hjust=-0.3) +
  labs(title = "Scatterplot of Life Expectancy for China against Canada",
       x = "Life Expectancy for China",
       y = "Life Expectancy for Canada") +
  theme_bw()
```

### 3. Re-lengthen

Re-lengthen the data.

```{r}
gapminder1 %>% 
  pivot_longer(cols = c(2:3),
               names_to = "country",
               names_prefix = "Life Expectancy for ",
               values_to = "lifeExp") %>% 
  arrange(country) %>% 
  DT::datatable()
```


## Exercise 2: Multivariate Option 1

### 1. Wider Format

Make a tibble with one row per year, and columns for life expectancy and GDP per capita for China and Canada.

```{r}
gapminder2 <- gapminder %>%
  filter(country %in% c("China","Canada")) %>% 
  select(year, country, lifeExp, gdpPercap) %>% 
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = c(lifeExp, gdpPercap)) 

DT::datatable(gapminder2)
```

### 2. Re-lengthen

Re-lengthen the data.

```{r,fig.pos="center"}
gapminder2 %>%
  pivot_longer(cols = c(2:5),
               names_to = c(".value", "country"),
               names_sep = "_") %>% 
  arrange(country) %>% 
  DT::datatable()
```

## Exercise 3: Table Joins

Read in the made-up wedding guestlist and email addresses.

```{r, message = FALSE}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```

### 3.1 Add Emails

For each guest in the guestlist (`guest` tibble), add a column for email address, which can be found in the `email` tibble.

```{r}
email1 <- email %>%
  separate_rows(guest, sep = ", ") %>% 
  rename(name = guest)

guest %>%
  left_join(email1, by = "name") %>%
  DT::datatable()
```

### 3.2 Uninvited Emails.

We have emails for Turner Jones, Turner Jones and Turner Jones, but they are not on the guestlist yet.

```{r}
email1 %>%
  anti_join(guest, by = "name") %>%
  DT::datatable()
```

### 3.3 Full guestlist.

Make a guestlist that includes everyone we have emails for (in addition to those on the original guestlist).

```{r}
guest %>% 
  full_join(email1, by = "name") %>% 
  DT::datatable()
```

