---
title: "Gapminder Exploration Presentation"
author: "Sihaoyu Gao"
institute: "The University of British Columbia"
date: "Sep 11, 2019"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, echo=FALSE,warning=FALSE}
options(width = 10)
knitr::opts_chunk$set(echo = TRUE, comment=NA, size = "small")
options(knitr.kable.NA = '')
```


```{r load, warning = FALSE, echo = FALSE}
library(gapminder)
library(ggplot2)
library(nlme)
library(knitr)
library(kableExtra)
```

# Introduction

This document explores the gapminder dataset. First of all, load the gapminder data set from `gapminder` library. Before doing any analysis, we need to understand the dataset. The data set is the excerpt of the Gapminder data on life expectancy, GDP per capita, and population by country. The main data frame gapminder has 1704 rows and 6 variables: 

```{r, eval=FALSE}
names(gapminder)
```

<br>

- country is a factor with 142 levels.
- continent is a factor with 5 levels.
- year ranges from 1952 to 2007 in increments of 5 years.
- lifeExp is life expectancy at birth (in years).
- pop is population, and gdpPercap is GDP per capita in US dollars.


---

# GDP per capita change for countries

Figure 1 shows the trajectories of GDP per capita over time for each country. There is one country had an extremely high GDP per capita before 1980, which is Kuwait in Asia.


```{r figs, echo=FALSE,fig.height=5.5,fig.width=8, fig.align='center'}
long.data<-groupedData(gdpPercap~year|country,data=gapminder) 
plot <- ggplot(data = gapminder, aes(x =year, y = gdpPercap, group =country))
plot + geom_line()+ geom_point()+labs(caption = "Figure 1: trajectories of GDP per capita over time for each country")+ theme(
  plot.caption = element_text(hjust = 0.5)
  )
```

---

# GDP per capita change for countries

- Singapore in Asia had the greatest increase in GDP per capital over 55 years, followed by Norway in Europe. 
- Kuwait in Asia had the least increase in GDP per capital, followed by Haiti in Americas. 


<br>
   
```{r,echo=FALSE, results='asis'}
data2<-gapminder[which(gapminder$year==2007),]
names(data2)[6] <- "GDP after"
data3<-gapminder[which(gapminder$year==1952),]
names(data3)[6] <- "GDP before"
data3<-data3[c(1,6)]
data4<-merge(data2,data3,by="country")
data4$change<-data4$`GDP after`-data4$`GDP before`
min<-data4[which(data4$change==min(data4$change)),]
min<-min[c(1,2,8)]
names(min)[3] <- "GDP change"
min<-cbind("least increase", min)
data4<-data4[which(data4$country != "Kuwait"),]
min2<-data4[which(data4$change==min(data4$change)),]
min2<-min2[c(1,2,8)]
names(min2)[3] <- "GDP change"
min2<-cbind(NA, min2)
names(min)[1] <- NA
names(min2)[1] <- NA
min<-rbind(min,min2)
row.names(min) <- NULL
max<-data4[which(data4$change==max(data4$change)),]
max<-max[c(1,2,8)]
names(max)[3] <- "GDP change"
max<-cbind("most increase", max)
data4<-data4[which(data4$country != "Singapore"),]
max2<-data4[which(data4$change==max(data4$change)),]
max2<-max2[c(1,2,8)]
names(max2)[3] <- "GDP change"
max2<-cbind(NA, max2)
names(max)[1] <- NA
names(max2)[1] <- NA
max<-rbind(max,max2)
row.names(max) <- NULL
table<-rbind(max,min)
names(table)[1] <- "Comment"
kable(table,caption = "Table 1: Countries with the most and the least increase in GDP per capita",align='c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))%>%
  column_spec(1,width = "3in") %>%
  column_spec(2,width = "3in") %>% column_spec(3,width = "3in") %>%
  column_spec(4,width = "3in") 
```


---

# GDP per capita change for continents

- Europe and Oceania had the fastest increase rate of GDP per capita.
- Africa had the lowest increase rate of GDP per capita over years. 

```{r fig_2, echo=FALSE,fig.align='center',fig.height=6,fig.width=9}
Africa<-gapminder[which(gapminder$continent=="Africa"),]
Americas<-gapminder[which(gapminder$continent=="Americas"),]
Asia<-gapminder[which(gapminder$continent=="Asia"),]
Europe<-gapminder[which(gapminder$continent=="Europe"),]
Oceania<-gapminder[which(gapminder$continent=="Oceania"),]
Africa<-aggregate(Africa[, 6], list(Africa$year), mean)
Africa<-cbind(Africa,continent="Africa")
America<-aggregate(Americas[, 6], list(Americas$year), mean)
America<-cbind(America,continent="America")
Asia<-aggregate(Asia[, 6], list(Asia$year), mean)
Asia<-cbind(Asia,continent="Asia")
Europe<-aggregate(Europe[, 6], list(Europe$year), mean)
Europe<-cbind(Europe,continent="Europe")
Oceania<-aggregate(Oceania[, 6], list(Oceania$year), mean)
Oceania<-cbind(Oceania,continent="Oceania")
newdata<-rbind(Africa,America,Asia,Europe,Oceania)
names(newdata)[1] <- "year"
plot <- ggplot(data = newdata, aes(x =year, y = gdpPercap, group =continent))
plot + geom_line(aes(colour=continent))+ geom_point()+labs(caption = "Figure 2: trajectories of GDP per capita over time for each continent")+ theme(
  plot.caption = element_text(hjust = 0.5)
  )
```

---

# GDP per capita vs. life expectancy

Life expectancy and GDP per capita have a positive correlation. 

```{r scatter, echo=FALSE, fig.align='center',fig.height=6,fig.width=7}
attach(gapminder)
plot(log(gdpPercap),lifeExp,pch=20, sub = "Figure 3: Relationship between GDP per capita (in log) and life expectancy")
```

---
# GDP per capita vs. life expectancy

The correlation coefficient also indicates the same result. 0.8 is quite close to 1, and it illustrates that life expectancy and GDP per capita (in log) have a positive correlation.

```{r}
cor(log(gdpPercap),lifeExp)
```


